---
title: "Visualizing Time Segmentation of Space"
author: "Josh Cullen"
date: "September 23, 2019"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Background

After running the initial time segmentation model on snail kite visitation to different grid cells (at 5 km resolution), the output is difficult to interpret with regard to underlying spatial patterns. To gain a better understanding of these patterns, segments will be plotted in multiple ways to discern how the time series has been partitioned.

The first plot will show the standard heatmap that we have initially used to plot mean changes in location and the calculated breakpoints. Next, the centroids of grid cells that were visited by each individual will be used to plot connections between relocations on a map over time. This will then be broken down further into centers of attraction (COAs) and peripheral locations. A 3D plot of space use over time will also be demonstrated for another perspective.

## Heatmaps
```{r wd and packages, warning=FALSE, message=FALSE}
set.seed(1)

library(dplyr)
library(lubridate)
library(ggplot2)
library(coda)
library(raster)
library(rnaturalearth)
library(rnaturalearthdata)

setwd("~/Documents/Snail Kite Project/Cat-Dirichlet Gibbs Sampler")
source('gibbs functions2.R')
```


```{r data import}
#Import data

setwd("~/Documents/Snail Kite Project/Data")
dat<- read.csv("Snail Kite Gridded Data.csv", header = T, sep = ",")
obs<- read.csv("Occupancy Matrix for all Obs and Locs.csv", header = T, sep = ",")
obs1.breakpts<- read.csv("ID1 Breakpoints (5 km).csv", header = T, sep = ",")
obs1.breakpts=obs1.breakpts[,1]
obs12.breakpts<- read.csv("ID12 Breakpoints (5 km).csv", header = T, sep = ",")
obs12.breakpts=obs12.breakpts[,1]
obs19.breakpts<- read.csv("ID19 Breakpoints (5 km).csv", header = T, sep = ",")
obs19.breakpts=obs19.breakpts[,1]
obs27.breakpts<- read.csv("ID27 Breakpoints (5 km).csv", header = T, sep = ",")
obs27.breakpts=obs27.breakpts[,1]


obs1=data.frame(obs) %>% filter(id == 1) %>% dplyr::select(-id)
obs12=data.frame(obs) %>% filter(id == 12) %>% dplyr::select(-id)
obs19=data.frame(obs) %>% filter(id == 19) %>% dplyr::select(-id)
obs27=data.frame(obs) %>% filter(id == 27) %>% dplyr::select(-id)

```


```{r plot segments 1, fig.cap="Heatmap of grid cell use over time in ID 1. Breakpoints calculated by model are denoted by vertical blue dashed lines."}
image(as.matrix(obs1)); title(x="Observations (n=9134)", y="Locations (n=378)", main="ID 1 (49 time segments)")
abline(v=obs1.breakpts/nrow(obs1),lty=3,col='blue')
```

```{r plot segments 12, fig.cap="Heatmap of grid cell use over time in ID 12. Breakpoints calculated by model are denoted by vertical blue dashed lines."}
image(as.matrix(obs12)); title(x="Observations (n=4922)", y="Locations (n=378)", main="ID 12 (48 time segments)")
abline(v=obs12.breakpts/nrow(obs12),lty=3,col='blue')
```

```{r plot segments 19, fig.cap="Heatmap of grid cell use over time in ID 19. Breakpoints calculated by model are denoted by vertical blue dashed lines."}
image(as.matrix(obs19)); title(x="Observations (n=3068)", y="Locations (n=378)", main="ID 19 (36 time segments)")
abline(v=obs19.breakpts/nrow(obs19),lty=3,col='blue')
```

```{r plot segments 27, fig.cap="Heatmap of grid cell use over time in ID 27. Breakpoints calculated by model are denoted by vertical blue dashed lines."}
image(as.matrix(obs27)); title(x="Observations (n=1182)", y="Locations (n=378)", main="ID 27 (8 time segments)")
abline(v=obs27.breakpts/nrow(obs27),lty=3,col='blue')
```





```{r assign time seg}

obs1$time1=1:nrow(obs1)
obs12$time1=1:nrow(obs12)
obs19$time1=1:nrow(obs19)
obs27$time1=1:nrow(obs27)

nloc=ncol(obs)-1 #remove time1

obs1.seg=get.summary.stats(obs1.breakpts,obs1,nloc)
obs12.seg=get.summary.stats(obs12.breakpts,obs12,nloc)
obs19.seg=get.summary.stats(obs19.breakpts,obs19,nloc)
obs27.seg=get.summary.stats(obs27.breakpts,obs27,nloc)



assign.time.seg<- function(obs, breakpts, dat) {
 seg.n=apply(obs,1,sum)
 time.seg=rep(1:(length(breakpts)+1), seg.n)
 dat$time.seg<- time.seg
 
 dat
}

dat1=dat %>% filter(id==1)
dat12=dat %>% filter(id==12)
dat19=dat %>% filter(id==19)
dat27=dat %>% filter(id==27)

dat1<- assign.time.seg(obs1.seg, obs1.breakpts, dat1)
dat12<- assign.time.seg(obs12.seg, obs12.breakpts, dat12)
dat19<- assign.time.seg(obs19.seg, obs19.breakpts, dat19)
dat27<- assign.time.seg(obs27.seg, obs27.breakpts, dat27)

```

## Spatial Maps

```{r map 1, fig.cap="Time segments are shown on continuous scale, which is makes plotting easier for large number of time segments."}
#Load world map data
world <- ne_countries(scale = "medium", returnclass = "sf")
usa<- world %>% filter(admin == "United States of America")


#Paths of individual segments ID 1
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat1, aes(x = utmlong, y = utmlat, color = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title="ID 1") +
  theme_bw()

#Paths of individual segments ID 12
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat12, aes(x = utmlong, y = utmlat, color = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title="ID 12") +
  theme_bw()

#Paths of individual segments ID 19
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat19, aes(x = utmlong, y = utmlat, color = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title="ID 19") +
  theme_bw()

#Paths of individual segments ID 1
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat27, aes(x = utmlong, y = utmlat, color = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title="ID 27") +
  theme_bw()
```


To view the differences among time segments from IDs with a large number of breakpoints (**IDs 1, 12, 19**), this is also plotted using only the first 10 time segments for better differentiation. As can be gleaned from this comparison, there is a large variety of movement patterns exhibited within only this short time period, therefore highlighting inter-individual differences in behavior.

```{r}
#Paths for segments ID 1
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat1[as.numeric(dat1$time.seg) <= 10,], aes(x = utmlong, y = utmlat,
                                                               color  = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title = "ID 1") +
  theme_bw()

#Paths for segments ID 12
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat12[as.numeric(dat12$time.seg) <= 10,], aes(x = utmlong, y = utmlat,
                                                               color  = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title = "ID 12") +
  theme_bw()

#Paths for segments ID 19
ggplot(data = usa) +
  geom_sf() +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)), ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat19[as.numeric(dat19$time.seg) <= 10,], aes(x = utmlong, y = utmlat,
                                                               color  = time.seg)) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Longitude", y = "Latitude", title = "ID 19") +
  theme_bw()
```

```{r create grid, message=FALSE}
# 5 km w 1 cell buffer
grid_5<- raster(extent(min(dat$utmlong), max(dat$utmlong),
                       min(dat$utmlat), max(dat$utmlat)) + 10000)
res(grid_5)<- 5000
proj4string(grid_5)<- CRS("+init=epsg:32617")
grid_5[]<- 0

#calc centroids of cells
grid.cell.locs<- coordinates(grid_5) %>% data.frame()

#Create grid cell borders
borders_5<- rasterToPolygons(grid_5, dissolve = F)
borders_5f<- fortify(borders_5)

#Make grid as DF
grid_5f<- as.data.frame(grid_5, xy = TRUE)

names(grid.cell.locs)<- c("grid.x", "grid.y")
grid.cell.locs$grid.cell<- 1:length(grid_5)

dat1<- left_join(dat1,grid.cell.locs, by="grid.cell")
dat12<- left_join(dat12,grid.cell.locs, by="grid.cell")
dat19<- left_join(dat19,grid.cell.locs, by="grid.cell")
dat27<- left_join(dat27,grid.cell.locs, by="grid.cell")
```

```{r grid map, message=FALSE}
#Plot time segments by grid cell ID 1
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_tile(data=grid_5f, aes(x=x, y=y), fill="transparent") +
  geom_path(data = borders_5f, aes(x=long, y=lat, group=group), size=0.25) +
  geom_point(data = dat1, aes(grid.x, grid.y, color=time.seg), size=1) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 1")

#Plot time segments by grid cell ID 12
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_tile(data=grid_5f, aes(x=x, y=y), fill="transparent") +
  geom_path(data = borders_5f, aes(x=long, y=lat, group=group), size=0.25) +
  geom_point(data = dat12, aes(grid.x, grid.y, color=time.seg), size=1) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 12")

#Plot time segments by grid cell ID 19
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_tile(data=grid_5f, aes(x=x, y=y), fill="transparent") +
  geom_path(data = borders_5f, aes(x=long, y=lat, group=group), size=0.25) +
  geom_point(data = dat19, aes(grid.x, grid.y, color=time.seg), size=1) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 19")

#Plot time segments by grid cell ID 27
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_tile(data=grid_5f, aes(x=x, y=y), fill="transparent") +
  geom_path(data = borders_5f, aes(x=long, y=lat, group=group), size=0.25) +
  geom_point(data = dat27, aes(grid.x, grid.y, color=time.seg), size=1) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 27")
```




## Centers of Attraction

To improve upon the previous set of maps, the next step is to determine the location(s) used within each time segment and accentuate which cells/locations are used the most. To support this visualization, locations that are used to a lesser extent will also be included to show the full extent of areas visited. By scaling the point size proportionally at each location per individual, these maps seek to relay a quick understanding of the primary relationship between the observed snail kite movements with space and time.

```{r}
mode.fun<- function(x) {
  names(table(x))[table(x)==max(table(x))]
  }

attr.cntrs1<- matrix(NA, length(unique(dat1$time.seg)), 4)
attr.cntrs12<- matrix(NA, length(unique(dat12$time.seg)), 4)
attr.cntrs19<- matrix(NA, length(unique(dat19$time.seg)), 4)
attr.cntrs27<- matrix(NA, length(unique(dat27$time.seg)), 4)


id.attr.cntr=function(dat,attr.cntrs) {
  for (i in 1:length(unique(dat$time.seg))) {
    tmp=mode.fun(dat[dat$time.seg == i, "grid.cell"])
    tmp=ifelse(length(tmp) > 1, tmp[1], tmp) #arbitrary selection of 1st cell in set if multiple vals
  
    attr.cntrs[i,1]=as.numeric(tmp)
    attr.cntrs[i,2]=grid.cell.locs[grid.cell.locs$grid.cell==attr.cntrs[i,1], "grid.x"]
    attr.cntrs[i,3]=grid.cell.locs[grid.cell.locs$grid.cell==attr.cntrs[i,1], "grid.y"]
    attr.cntrs[i,4]=max(table(dat[dat$time.seg == i, "grid.cell"]))
  }
  colnames(attr.cntrs)<- c("grid.cell","grid.x","grid.y","size")
  attr.cntrs<- data.frame(attr.cntrs)
  attr.cntrs$time.seg<- 1:length(unique(dat$time.seg))
  
  attr.cntrs
}


attr.cntrs1=id.attr.cntr(dat1, attr.cntrs1)
attr.cntrs12=id.attr.cntr(dat12, attr.cntrs12)
attr.cntrs19=id.attr.cntr(dat19, attr.cntrs19)
attr.cntrs27=id.attr.cntr(dat27, attr.cntrs27)
```

```{r map COA, message=FALSE}
#Map of connected COA for ID 1
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = attr.cntrs1, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs1, aes(x=grid.x, y=grid.y, color = time.seg), size=log10(attr.cntrs1$size)*2) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 1") +
  theme_bw()

#Map of connected COA for ID 12
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = attr.cntrs12, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs12, aes(x=grid.x, y=grid.y, color = time.seg), size=log10(attr.cntrs12$size)*2) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 12") +
  theme_bw()

#Map of connected COA for ID 19
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = attr.cntrs19, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs19, aes(x=grid.x, y=grid.y, color = time.seg), size=log10(attr.cntrs19$size)*2) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 19") +
  theme_bw()

#Map of connected COA for ID 27
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = attr.cntrs27, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs27, aes(x=grid.x, y=grid.y, color = time.seg), size=log10(attr.cntrs27$size)*2) +
  scale_color_viridis_c("Time Segment") +
  labs(x = "Easting", y = "Northing", title = "ID 27") +
  theme_bw()
```

```{r scale point size}
dat.size1=matrix(NA,nrow(dat1),1)
dat.size12=matrix(NA,nrow(dat12),1)
dat.size19=matrix(NA,nrow(dat19),1)
dat.size27=matrix(NA,nrow(dat27),1)

n.per.tseg=function(dat,dat.size) {
  for (i in 1:length(unique(dat$time.seg))) {
    tmp=table(dat[dat$time.seg == i, "grid.cell"]) %>% data.frame()
    ind=which(dat$time.seg == i)
    for (j in ind){
      dat.size[j,]=tmp[which(tmp$Var1 == dat$grid.cell[j]),2]
    }
  }
  dat.size
}

dat1$size<- n.per.tseg(dat1, dat.size1) %>% as.vector()
dat12$size<- n.per.tseg(dat12, dat.size12) %>% as.vector()
dat19$size<- n.per.tseg(dat19, dat.size19) %>% as.vector()
dat27$size<- n.per.tseg(dat27, dat.size27) %>% as.vector()
```

```{r map all scaled points, message=FALSE}
#Map of scaled pts ID 1
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat1, aes(x=grid.x, y=grid.y), color="gray60", size=0.5) +
  geom_point(data = dat1, aes(x=grid.x, y=grid.y, color = as.numeric(time.seg)),
             size=log10(dat1$size)*2) +
  geom_path(data = attr.cntrs1, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs1, aes(x=grid.x, y=grid.y, color = time.seg),
             size=log10(attr.cntrs1$size)*2) +
  scale_color_viridis_c("Time Segment (n=49)") +
  labs(x = "Easting", y = "Northing", title = "ID 1") +
  theme_bw()

#Map of scaled pts ID 12
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat12, aes(x=grid.x, y=grid.y), color="gray60", size=0.5) +
  geom_point(data = dat12, aes(x=grid.x, y=grid.y, color = as.numeric(time.seg)),
             size=log10(dat12$size)*2) +
  geom_path(data = attr.cntrs12, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs12, aes(x=grid.x, y=grid.y, color = time.seg),
             size=log10(attr.cntrs12$size)*2) +
  scale_color_viridis_c("Time Segment (n=48)") +
  labs(x = "Easting", y = "Northing", title = "ID 12") +
  theme_bw()

#Map of scaled pts ID 19
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat19, aes(x=grid.x, y=grid.y), color="gray60", size=0.5) +
  geom_point(data = dat19, aes(x=grid.x, y=grid.y, color = as.numeric(time.seg)),
             size=log10(dat19$size)*2) +
  geom_path(data = attr.cntrs19, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs19, aes(x=grid.x, y=grid.y, color = time.seg),
             size=log10(attr.cntrs19$size)*2) +
  scale_color_viridis_c("Time Segment (n=36)") +
  labs(x = "Easting", y = "Northing", title = "ID 19") +
  theme_bw()

#Map of scaled pts ID 27
ggplot() +
  geom_sf(data = usa) +
  coord_sf(datum = NA) +
  coord_sf(crs = "+init=epsg:32617", xlim = c(min(dat$utmlong-20000), max(dat$utmlong+20000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat27, aes(x=grid.x, y=grid.y), color="gray60", size=0.5) +
  geom_point(data = dat27, aes(x=grid.x, y=grid.y, color = as.numeric(time.seg)),
             size=log10(dat27$size)*2) +
  geom_path(data = attr.cntrs27, aes(x=grid.x, y=grid.y)) +
  geom_point(data = attr.cntrs27, aes(x=grid.x, y=grid.y, color = time.seg),
             size=log10(attr.cntrs27$size)*2) +
  scale_color_viridis_c("Time Segment (n=8)") +
  labs(x = "Easting", y = "Northing", title = "ID 27") +
  theme_bw()
```




## 3D Plot

Another potential method to view these time series data is by a 3D scatter plot. Since the static version of these plots is not as useful as the dynamic/interactive form, the online hosted version of these plots for [ID 1](https://plot.ly/~joshcullen/28), [ID 12](https://plot.ly/~joshcullen/30), [ID 19](https://plot.ly/~joshcullen/32), and [ID 27](https://plot.ly/~joshcullen/34) have been included.

